---
title: "HW3"
format: html
editor: visual
---

### Github URL: https://github.com/aelbass/STATS506_HW1.git

## Problem 1

### a. Read data

```{r}
#installed 'haven' in console
library(haven)

data <- read_xpt("AUX_I.xpt")
data2 <- read_xpt("DEMO_I.xpt")

#Merge using (dplyr)

#installed 'dplyr' in console
library(dplyr)

merged <- inner_join(data, data2, by = "SEQN")
dim(merged)
```


### b. Clean up data


```{r}
#Clean up gender
unique(merged$RIAGENDR)
merged$RIAGENDR <- factor(merged$RIAGENDR,
                      levels = c(1, 2),
                      labels = c("male", "female"))
unique(merged$RIAGENDR)
```

Gender data is already clean; I just categorized to male and female.


```{r}
#Clean up citizenship

merged$DMDCITZN[!merged$DMDCITZN %in% c(1, 2, 7, 9)] <- NA

merged$DMDCITZN <- factor(merged$DMDCITZN,
                          levels = c(1, 2, 7, 9),
                          labels = c("citizen", "non-citizen", "refused", "don't know"))

library(forcats)
merged$DMDCITZN <- fct_explicit_na(merged$DMDCITZN, na_level = "missing")

unique(merged$DMDCITZN)
table(merged$DMDCITZN)
```


Cleaned up citizenship status and turned into factor with categorical levels.


```{r}
#Clean up children under the age of 5 data
unique(merged$DMDHHSZA)
```

Children under the age of 5 data is already clean and not categorical.


```{r}
#Clean up annual household income
unique(merged$INDHHIN2)
library(dplyr)

merged <- merged %>%
  mutate(INDHHIN2 = case_when(
    INDHHIN2 == 12 ~ 14,
    INDHHIN2 == 14 ~ 11,
    INDHHIN2 == 15 ~ 12,
    TRUE ~ INDHHIN2  # keep everything else unchanged
  ))
table(merged$INDHHIN2)
```


I changed up the ordering of the categories: I made the "0 to 4,999", "5,000 to 9,999", etc. take 1-12 (including 11) for code or value, "Under $20,000" take 13, "20,000 and Over" take 14. There was no missing data.


### c. Poisson regression models predicting a respondent’s Tympanometric width in each ear


```{r}

```


Spare

```{r}

```