---
title: "HW4"
format: html
editor: visual
---

### Github URL: https://github.com/aelbass/STATS506_HW1.git

## Problem 1

### a. Generate a table (which can just be a nicely printed tibble) of vote count (regardless of party) per year/type. Make sure to sort it by vote count.

```{r}
library(nzelect)
library(tidyverse)
library(tidyr)
library(dplyr)

vote_summary <- nzge %>%
  group_by(election_year, voting_type) %>%
  summarise(total_votes = sum(votes, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_votes))

vote_summary
```


### b. Focus only on the 2014 election. Report the proportion of votes for each party in the Candidate election. Again, produce a nice table and sort it by percent of vote.


```{r}
candidate_2014 <- nzge %>%
  filter(election_year == 2014, voting_type == "Candidate") %>%
  group_by(party) %>%
  summarise(total_votes = sum(votes, na.rm = TRUE), .groups = "drop") %>%
  mutate(percent_vote = round(100 * total_votes / sum(total_votes), 2)) %>%
  select(party, percent_vote) %>% 
  arrange(desc(percent_vote))

candidate_2014
```

###c. Produce a nice table indicating, for each year, which party won the Candidate vote and which party won the Party vote.

```{r}
winners_by_year <- nzge %>%
  group_by(election_year, voting_type, party) %>%
  summarise(total_votes = sum(votes, na.rm = TRUE), .groups = "drop") %>%
  group_by(election_year, voting_type) %>%
  slice_max(total_votes, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  
  
  pivot_wider(names_from = voting_type,
              values_from = party,
              names_prefix = "winner_") %>%
  arrange(election_year)

winners_by_year
```

## Problem 2


```{r}
atp_matches_2019 <- read_csv("atp_matches_2019.csv")
head(atp_matches_2019)
glimpse(atp_matches_2019)
```

### a. How many tournaments took place in 2019

```{r}
atp_matches_2019 %>%
  mutate(year = str_extract(tourney_id, "^[0-9]{4}")) %>%
  filter(year == "2019") %>%
  summarise(tournaments_2019 = n_distinct(tourney_id))
```

128 tournaments took place in 2019.


### b. Did any player win more than one tournament? If so, how many players won more than one tournament, and how many tournaments did the most winning player(s) win?

```{r}
tournament_winners <- atp_matches_2019 %>%
  group_by(tourney_id) %>%
  filter(match_num == max(match_num, na.rm = TRUE)) %>%  # final match per tournament
  ungroup() %>%
  select(tourney_id, winner_name)

player_wins <- tournament_winners %>%
  count(winner_name, name = "tournaments_won") %>%
  arrange(desc(tournaments_won))
players_more_than_one <- player_wins %>%
  filter(tournaments_won > 1)

num_players_more_than_one <- nrow(players_more_than_one)
max_tournaments_won <- max(player_wins$tournaments_won)

player_wins
num_players_more_than_one
max_tournaments_won
```

From the table above, 17 players won more than one tournament. The most winning player won 9 tournaments, and that was Rafael Nadal.


### c. Is there any evidence that winners have more aces than losers? (If you address this with a hypothesis test, do not use base R functionality - continue to remain in the Tidyverse.)

```{r}
aces_data <- atp_matches_2019 %>%
  select(w_ace, l_ace) %>%
  pivot_longer(cols = c(w_ace, l_ace),
               names_to = "player_type",
               values_to = "aces") %>%
  mutate(player_type = if_else(player_type == "w_ace", "Winner", "Loser"))

aces_data %>%
  group_by(player_type) %>%
  summarise(mean_aces = mean(aces, na.rm = TRUE),
            median_aces = median(aces, na.rm = TRUE),
            sd_aces = sd(aces, na.rm = TRUE),
            n = n())


```

Winners tend to average more aces than losers.

I am not sure if it is enough to get average and std, so I ran a hypothesis test below with a null hypothesis that there is no difference in the mean number of aces between winners and losers.

```{r}
library(tidyverse)
library(broom)

ace_tidytest <- aces_data %>%
  group_by(player_type) %>%
  summarise(aces = list(aces)) %>%
  summarise(tidy_result = list(tidy(t.test(aces[[1]], aces[[2]], var.equal = FALSE)))) %>%
  unnest(tidy_result)

ace_tidytest
```

And since p < 0.05, there’s evidence that winners tend to serve more aces.


### d. Identify the player(s) with the highest win-rate. (Note that this is NOT asking for the highest number of wins.) Restrict to players with at least 5 matches.

```{r}
player_stats <- atp_matches_2019 %>%
  select(winner_name, loser_name) %>%
  pivot_longer(cols = c(winner_name, loser_name),
               names_to = "result",
               values_to = "player") %>%
  mutate(is_win = if_else(result == "winner_name", 1, 0))

player_summary <- player_stats %>%
  group_by(player) %>%
  summarise(
    matches = n(),
    wins = sum(is_win),
    win_rate = wins / matches
  ) %>%
  filter(matches >= 5) %>%
  arrange(desc(win_rate))

player_summary %>%
  filter(win_rate == max(win_rate))
```

Rafael Nadal is has the highest win rate, ~87%. He played 69 games.


Template

```{r}

```


