---
title: "HW6"
format: html
editor: visual
---

### Github URL: https://github.com/aelbass/STATS506_HW1.git

## Problem 1

```{r}
library(e1071)
library(Rcpp)

cppFunction("
double C_moment(NumericVector v, int k) {
  int n = v.size();
  double mean = 0.0;

  for (int i = 0; i < n; ++i) {
    mean += v[i];
  }
  mean /= n;

  // compute k-th central moment
  double total = 0.0;
  for (int i = 0; i < n; ++i) {
    total += pow(v[i] - mean, k);
  }

  return total / n;
}
")

set.seed(7)
x <- rnorm(1000)

C_moment(x, 2)
moment(x, order = 2, center = TRUE)

C_moment(x, 3)
moment(x, order = 3, center = TRUE)

C_moment(x, 4)
moment(x, order = 4, center = TRUE)
```

## Problem 4

### a. How many tournaments took place in 2019

```{r}
library(data.table)

atp <- fread("atp_matches_2019.csv")

atp[, year := substr(tourney_id, 1, 4)
    ][year == "2019",
      .(tournaments_2019 = uniqueN(tourney_id))]
```

### b. Did any player win more than one tournament? If so, how many players won more than one tournament, and how many tournaments did the most winning player(s) win?

```{r}
tournament_winners <- atp[
  , .SD[which.max(match_num)],
  by = tourney_id
][
  , .(tourney_id, winner_name)
]

player_wins <- tournament_winners[
  , .(tournaments_won = .N),
  by = winner_name
][
  order(-tournaments_won)
]

players_more_than_one <- player_wins[tournaments_won > 1]
num_players_more_than_one <- nrow(players_more_than_one)
max_tournaments_won <- max(player_wins$tournaments_won)

num_players_more_than_one
max_tournaments_won
```

Yes, 17 players won more than one tournament. The most winning player won 9 tournaments, and that was Rafael Nadal.

### c. Is there any evidence that winners have more aces than losers? (If you address this with a hypothesis test, do not use base R functionality - continue to remain in the Tidyverse.)

```{r}
aces_data <- melt(
  atp,
  measure.vars = c("w_ace", "l_ace"),
  variable.name = "player_type",
  value.name = "aces"
)[
  , player_type := fifelse(player_type == "w_ace", "Winner", "Loser")
]

# 2. Summary statistics by player_type
aces_summary <- aces_data[
  , .(
      mean_aces   = mean(aces, na.rm = TRUE),
      median_aces = median(aces, na.rm = TRUE),
      sd_aces     = sd(aces, na.rm = TRUE),
      n           = .N
    ),
  by = player_type
]

aces_summary
```

Winners tend to average more aces than losers. Also here is a null hypothesis test that there is no difference in the mean number of aces between winners and losers.

```{r}
library(broom)

DT <- as.data.table(aces_data)

ace_list <- DT[
  , .(aces_list = list(aces)),
  by = player_type
]

tt <- t.test(
  ace_list[player_type == "Winner", aces_list][[1]],
  ace_list[player_type == "Loser", aces_list][[1]],
  var.equal = FALSE
)

ace_tidytest <- tidy(tt)

ace_tidytest

```

And since p \< 0.05, there’s evidence that winners tend to serve more aces.


### d. Identify the player(s) with the highest win-rate. Restrict to players with at least 5 matches.


```{r}

DT1 <- as.data.table(atp)

player_stats <- melt(
  DT1,
  measure.vars = c("winner_name", "loser_name"),
  variable.name = "result",
  value.name = "player"
)[

  , is_win := fifelse(result == "winner_name", 1L, 0L)
]

player_summary <- player_stats[
  , .(
      matches  = .N,
      wins     = sum(is_win),
      win_rate = sum(is_win) / .N
    ),
  by = player
][
  matches >= 5
][
  order(-win_rate)
]

best_players <- player_summary[win_rate == max(win_rate)]

best_players
```

Rafael Nadal is has the highest win rate, \~87%. He played 69 games.
